Region-network hierarchical sparsity priors for\\ high-dimensional inference in brain imaging

### 2016-01-21 paper/paper.tex
intro+disc somewhat done

\textbf{\\keywords}:
Sparsity-inducing Norms, Structured Sparsity, Numerical Optimization,
Systems Neuroscience, Brain Imaging,
Functional Specialization, Functional Integration

@@

\section{Introduction}
% sparsity
Many quantitative scientific domains underwent a
recent passage from the classical regime (i.e., "long data")  to
the high-dimensional regime (i.e., "wide data")
\cite{jordan2015massive}.
Also in the brain imaging domain,
many contemporary methods for acquiring brain signals yield
more variables per observation than
total observations available in a data sample.
This scenario challenges many statistical estimators from
classical statistics.
For instance,
linear regression models without additional assumptions
yield an infinity of possible coefficients
and, thus, no solution.
%
\textit{Sparsity} assumptions have consequently been
introduced and made many ill-posed estimation problems tractable
\cite{buhlmann2011statistics, hastie2015statistical}.
Domain-specific structure can be imposed on the 
statistical estimation,
thus preassuming that variables have unequal importance
and follow existing knowledge
\cite{bach2012optimization}.
Sparsified supervised and unsupervised
learning algorithms have proven to yield
statistical relationships that can be readily
estimated, reproduced, and interpreted
\cite{giraud2014introduction}.
%
Yet, what neurobiological structure suggests itself
to successfully tackle the
\textit{curse of dimensionality} in
functional neuroimaging research?

% specialization & integration
Concepts on human brain organization have long been torn
between the two extremes
\textit{functional specialization} and \textit{functional integration}.
Functional specialization emphasizes that microscopically distinguishable
brain regions probably solve distinct classes of computational processes
\cite{kanwisher2010functional}.
Functional integration, in turn, emphasizes that brain function
is probably enabled by complex connections between these
distinct brain regions \cite{sporns14nn}.
%
These notions were predominantly derived from
invasive examination of \textit{structure} (i.e., histological preparation),
\textit{connectivity}, (i.e., axonal tracing),
and \textit{functional properties}
(i.e., single cell recordings) in same animals.
Regarding functional segregation into specialized regions,
early histological investigations into the microscopic heterogeneity of
the human cerebral cortex have resulted
in several detailed anatomical maps
\cite{brodmann1909vergleichende, vogt1919allgemeine}.
Regarding axonal connections,
each such cortical area has been observed
to possess a unique set of incoming and outgoing connections
\cite{passingham2002, young93monkey, scannell95cat}.
%
Both
local cyto- and chemoarchitectonic infrastructure
and its unique global connectivity profile
together realize computational processes (i.e., function).
While
cortical modules versus connections between those
reflect 
functional specialization versus functional integration
\cite{friston2002beyond, mesulam_sensation},
both architectural principles are conceptually inextricable when
explaining the emergence of mental operations
\cite{tononi1998complexity, saygin2012}.

% SPECIAL
Functional specialization has been
explored and interpreted based on many different research methods.
%
Single cell recordings and microscopic examination
revealed, for instance, the
specialization in the visual cortex into V1, V2, V3, V3A, and V4
\cite{hubel1962receptive, zeki1978functional}.
Tissue lesion of the mid-fusiform gyrus of the visual system
were frequently reported to impair
recognition of others' identity from faces
\cite{iaria2008contrib}.
The localization of
sensory, motor, and emotional cognitive functions to cortical areas
in the living brain
has later been enabled by
non-invasive brain imaging with
functional magnetic resonance imaging (fMRI) and
positron emission tomography (PET)
\cite{fristen1997imaging}.
Further,
radioactive mapping of neurotransmitter receptor distributions
rendered accessible yet another
local characteristic of neuronal populations
\cite{zilles2009receptor}.
% allude to our many-regions layer in our methods
In the computational era,
automatic clustering methods are increasingly employed to
regionally differentiate the cerebral cortex,
which can partly be more fine-grained than
classical cytoarchitectonic borders
\cite{behrens03, cbp2015review}.
Today,
high-thoughput computing enables 
ultahigh-resolution 3D models of brain anatomy
at macroscopical to near-cellular scale
\cite{amunts2013bigbrain}.
%
As a crucial common point,
all these methodological approaches
yield neuroscientific findings
that are naturally interpreted according to
non-overlapping, discrete region compartments
as the basic architectonic framework of brain organization.

% INTEGRAL
It is more recent
that the main interpretational focus has shifted
from circumscribed regions to network stratifications
in systems neuroscience \cite{yuste2015, stephan_dys}.
The higher the processing hierarchy level,
the more a cortical area is known to
connect to different large-scale networks
\cite{yeo11}.
%
Invasive axonal tracing studies in monkeys were complemented
by in-vivo diffusion MRI tractography in humans
as a now frequently employed method to
outline fiber bundles between brain regions
\cite{jbabdi2013long}.
Besides analyses of
electrophysiological oscillations
\cite{buzsaki2004neuronal}
and
graph-theoretical properties \cite{bullmore2009complex},
studies of
functional connectivity \cite{buckner2013opportunities} and
independent component analysis (ICA) \cite{beckmann2005}
became the workhorses of network discovery
in functional neuroimaging.
These revealed the important implication of
canonical brain networks across cognitive domains,
including the so-called
``default-mode network'' \cite{raichle2001pnas},
``salience network'' \cite{seeley2007dissociable},
and ``dorsal attention network'' \cite{corbettashul2008}. 
Characteristic changes in the configuration of
these large-scale networks
were repeatedly observed to be induced
by the onset of given cognitive tasks \cite{fransson2006}.
Such task-induced mechanisms orchastrating supraordinate networks
might be subserved by the right anterior insula \cite{sridh2008}
and temporo-parietal junction \cite{bzdok2013tpj}.
%
Ultimately,
interpretation of findings from all these methods naturally embraces
cross-regional integration by
overlapping network compartments
as the basic architectonic framework of brain organization,
in stark constrast to methods examining regional specialization.

% study
Building on these interpretational traditions,
the present study proposes to incorporate
known architectonic aspects from both
anatomical segregation and integration
into unified classification and regression algorithms
by means  of structured sparsity assumptions.
The "true" relative importance of 
local region compartments and global network compartments
is typically unknown
but
probably varying in degree
across diverse neuroscientific questions.
Recent advances in multivariate statistical learning techniques
enable such adaptive estimators with
neurobiologically plausible region and network priors.
Using a large reference dataset,
we demonstrated that domain-informed supervised models
gracefully handle the curse of dimensionality,
yield many more human-interpretable results,
and generalize better to new samples
than domain-na\"ive models.

higher-order prior knowledge

structured sparsity
extend trend in stastical modelling from
processing of auditory signals \cite{daudet2004sparse},
natural images \cite{harzallah2009combining} and
videos \cite{kang2015structured, kim2010sparse}
to
genetics \cite{rapaport2008classification, kim2012tree},
astrophysics \cite{vinci2014estimating},
and
conformational dynamics of protein complexes \cite{jenatton2009structured}.

tructured-sparsity-inducing penalty

to brain imaging domain.

@@

\section{Discussion}

exploit known structure for variable selection and prediction
in classification and regression

Among a collection of plau- sible models, the simplest one is often preferred

Functional specialization and functional integration
in the human brain are
probably both mechanistically relevant for emerging behavior.
However, there is a scarcity of statistical methods
that concomitantly relate brain regions and networks to
an organism's repertoire of mental operations.

However, in the problem of multi-task regression, where the output is a multivariate vector with an internal sparsity struc- ture, the estimation of the regression parameters can potentially benefit from taking into account this spar- sity structure in the estimation process. This will allow the strongly related output variables to be mapped to the input factors in a synergistic way, which is not possible in the standard lasso.

We assume regions and networks but can estimate 
importance of each voxel.

neurobiologically motivated restrictions to
complexity circumvented the curse of dimensionality and allowed
for useful, simplified views on brain function.

The same classification performance can be reached by
many different models -> we naturally want to priviledge those
among the equally performing models that are closest to
what we know about the brain

-> interpreta bility and accuracy typically exclude each other
(cf. Hastie)

alternatives would be dimensionality reduction
and dictionary learning

Among a collection of plausible models,
the simplest one should be preferred

curse of dimensionality has classically been circumvented
by treating a high-dimensional problem
as a mass-univariate problem.
Problem was postponed to the inference step
this is why we never solved the multiple comparisons problem in fMRI

regularization with sparsity penalties
restrictions to complexity according to neurobiological facts
-> bias-variance tradeoff: negigible increase in bias but large decrease in variance
-> VC: decrease complexitiy capactiy -> better chance for generalization out of sample

multiple comparisons problem has the curse of dimensionality
as its close correspondence in statistical learning
sparsity -> solves the curse at the estimation level

OUTLOOK:
robust high-dimensional inference will be key
to prediction in single individual,
such as for personalized medicine
\cite{gabrieli2015prediction}

More generally, our regularization scheme could also be used for various
learning tasks, as soon
as prior knowledge on the structure of the sparse representation
is available, e.g.,
for multiple kernel learning (Micchelli and Pontil, 2006),
multi-task learning (Argyriou et al., 2008; Obozinski et al., 2009;
Kim and Xing, 2009) and sparse matrix factorization
problems (Mairal et al., 2010; Jenatton et al., 2010).

this type of prior knowledge is of clear interest for the spatially and tempo-
rally structured data typical in bioinformatics, computer vision and neuroscience applications

structured sparsity can even be used for
unsupervised data modelling,
such as in PCA \cite{jenatton2009structured}

Differences to Bertrands structured sparsity paper:
- they derive the tree from the data, we use a priori knowledge from
the existing literature
-

>>>>

\textbf{\\keywords}:
Sparsity-inducing norms, hierarchical structured sparsity,
numerical optimization,
systems neuroscience, brain imaging,
functional specialization, functional integration

@@

\section{Introduction}
% sparsity
Many quantitative scientific domains underwent a
recent passage from the classical regime (i.e., "long data")  to
the high-dimensional regime (i.e., "wide data")
\cite{jordan2015massive}.
Also in the brain imaging domain,
many contemporary methods for acquiring brain signals yield
more variables per observation than
total observations per data sample.
This high-dimensional scenario challenges many statistical estimators from
classical statistics.
For instance,
the general linear model without additional assumptions
yields an infinity of possible coefficients
and, thus, no solution.
%
Many such ill-posed estimation problems
have become tractable by means of
\textit{sparsity} assumptions
\cite{buhlmann2011statistics, hastie2015statistical}.
Sparsified supervised and unsupervised
learning algorithms have proven to yield
statistical relationships that can be readily
estimated, reproduced, and interpreted
\cite{giraud2014introduction}.
%
Further, \textit{structured sparsity} can impose
domain knowledge on the 
statistical estimation,
thus preassuming the variables to have unequal importance
and to obey expected data distributions
\cite{bach2012optimization}.
Yet, what neurobiological structure suggests itself
to harness the
\textit{curse of dimensionality} with
$\textgreater$100,000 variables
in neuroimaging research?

% specialization & integration
Concepts on human brain organization have long been torn
between the two extremes
\textit{functional specialization} and \textit{functional integration}.
Functional specialization emphasizes that microscopically distinguishable
brain regions solve distinct classes of computational processes
\cite{kanwisher2010functional}.
Functional integration, in turn, emphasizes that brain function
is enabled by complex connections between these
distinct brain regions \cite{sporns14nn}.
%
These notions were predominantly derived from
invasive examination of anatomy (i.e., histological preparation),
connectivity, (i.e., axonal tracing),
and functional properties
(i.e., single cell recordings) in animals.
Regarding functional segregation into specialized regions,
early histological investigations into the microscopic heterogeneity of
the human cerebral cortex have resulted
in several detailed anatomical maps
\cite{brodmann1909vergleichende, vogt1919allgemeine}.
Regarding axonal connections,
each such cortical area has been observed
to possess a unique set of incoming and outgoing connections
\cite{passingham2002, young93monkey, scannell95cat}.
%
Both
local
%cyto- and chemoarchitectonic
infrastructure
and its unique global connectivity profile
together are thought to realize brain function.
%
In sum,
cortical brain modules versus connections between those
reflect 
functional specialization versus functional integration
\cite{friston2002beyond, mesulam_sensation}.
Importantly,
probably no existing brain analysis method acknowledges that
both architectural principles are inextricably involved
in the realization of mental operations
\cite{tononi1998complexity, saygin2012}.

% SPECIAL
Functional specialization has been
explored and interpreted based on many different research methods.
%
Single cell recordings and microscopic examination
revealed, for instance, the
specialization in the visual cortex into V1, V2, V3, V3A, and V4
\cite{hubel1962receptive, zeki1978functional}.
Tissue lesion of the mid-fusiform gyrus of the visual system,
for instance,
was frequently reported to impair
recognition of others' identity from faces
\cite{iaria2008contrib}.
The whole-brain localization of
sensory, motor, and emotional functions to cortical areas
% in the living brain
has later been enabled by
non-invasive brain imaging with
functional magnetic resonance imaging (fMRI) and
positron emission tomography (PET)
\cite{fristen1997imaging}.
Further,
radioactive mapping of neurotransmitter receptors
rendered accessible yet another
local characteristic of neuronal populations
\cite{zilles2009receptor}.
% allude to our many-regions layer in our methods
In the computational era,
automatic clustering methods are increasingly employed to
regionally differentiate the cerebral cortex,
which can partly be more fine-grained than
classical microscopical borders
\cite{behrens03, cbp2015review}.
Today,
high-thoughput approaches enable
ultahigh-resolution 3D models of brain anatomy
at near-cellular scale
\cite{amunts2013bigbrain}.
%
As a crucial common point,
all these methodological approaches
yield neuroscientific findings
that are naturally interpreted according to
non-overlapping, discrete region compartments
as the basic architecture of brain organization.

% INTEGRAL
It is more recent
that the main interpretational focus has shifted
from circumscribed regions to network stratifications
in systems neuroscience \cite{yuste2015, stephan_dys}.
%
Invasive axonal tracing studies in monkeys were complemented
by diffusion MRI tractography in humans
as a now frequently employed method to
outline fiber bundles between brain regions
\cite{jbabdi2013long}.
Besides analyses of
electrophysiological oscillations
\cite{buzsaki2004neuronal}
and
graph-theoretical properties \cite{bullmore2009complex},
studies of
functional connectivity \cite{buckner2013opportunities} and
independent component analysis (ICA) \cite{beckmann2005}
became the workhorses of network discovery
in neuroimaging.
These revealed the important implication of
canonical brain networks across cognitive domains,
including the so-called
``default-mode network'' \cite{raichle2001pnas},
``salience network'' \cite{seeley2007dissociable},
and ``dorsal attention network'' \cite{corbettashul2008}. 
Characteristic changes in the configuration of
these macroscopical networks
were repeatedly observed to be induced
by the onset of given cognitive tasks \cite{fransson2006}.
Such task-induced mechanisms orchastrating supraordinate networks
might be subserved by the right anterior insula \cite{sridh2008}
and temporo-parietal junction \cite{bzdok2013tpj}.
%
Ultimately,
interpretation of findings from all these methods naturally embraces
cross-regional integration by
overlapping network compartments
as the basic architecture of brain organization,
in stark constrast to methods examining regional specialization.

% study
Building on these two major interpretational streams in systems neuroscience,
the present study proposes to incorporate
established neurobiological structure underlying
functional segregation and integration
into supervised estimators
by hierarchical structured sparsity.
% The "true" relative importance of 
% local region compartments and global network compartments
% is typically unknown
% but
% probably varying in degree
% across diverse neuroscientific questions.
%
Learning techniques
exploiting structured sparsity 
have recently made much progress in various application domains
from processing of auditory signals \cite{daudet2004sparse},
natural images \cite{harzallah2009combining} and
videos \cite{kang2015structured, kim2010sparse}
to
genetics \cite{rapaport2008classification, kim2012tree},
astrophysics \cite{vinci2014estimating},
and
conformational dynamics of protein complexes \cite{jenatton2009structured}.
%
This is extended by the present work that
enables neuroscience-specific estimators capitalizing on
neurobiologically plausible region and network priors.
%
Using a large reference dataset,
we demonstrated that domain-informed supervised models
gracefully tackle the curse of dimensionality,
yield more human-interpretable results,
and generalize better to new samples
than domain-na\"ive estimators.

@@

\section{Discussion}
Relevant structure in brain imaging data has long been
separately investigated
along two distinct organizational principles:
functional segregation into discrete brain regions
and functional integration by inter-regional brain networks.
%
This paper demonstrates the simultaneous exploitation of
both these neurobiological compartments
for sparse variable selection and high-dimensional prediction.
%
Introducing existing domain knowledge into model selection
naturally priviledges model hypotheses in the function space
that are most neurobiologically plausible.
%
The neurobiologically informed
hierarchical structured sparsity penalty is shown to enhance
model interpretability and generalization performance,
although these statistical goals are typically in conflict.

1) each voxel
We assume regions and networks but can estimate 
importance of each voxel.
neurobiologically motivated restrictions to
complexity circumvented the curse of dimensionality and allowed
for useful, simplified views on brain function.
previously laborious feature engineering / selection procedures

2) restrictions to complexity according to neurobiological facts
-> bias-variance tradeoff: negigible increase in bias but large decrease in variance
-> VC: decrease complexitiy capactiy -> better chance for generalization out of sample

3)
multiple comparisons problem has the curse of dimensionality
as its close correspondence in statistical learning
sparsity -> solves the curse at the estimation level

4) region-constrained or network-constrained model estimation
can lead to similar predictive performance

In the future,
the proposed regularization scheme can be used for various
learning tasks in systems neuroscience.
This includes supervised classification and regression
with one or several target variables.
%
Region-network sparsity priors could even be used to inform
unsupervised statistical estimators,
including principal component analysis
\cite{jenatton2009structured}
and
k-means clustering \cite{???}.
%
Additionally,
the structured sparsity penalty could be extended
from the spatial domain of neural activity to
spatiotemporally coherent neural activity patterns
\cite{gramfort2011tracking}.
%
Ultimately,
successful high-dimensional inference is
an important prerequisite
for prediction of diagnosis,
disease trajectories, and treatment responses
for personalized medicine in psychiatry and neurology
\cite{gabrieli2015prediction}.
